Exploring dimensions of biodiversity in Japanese ferns

### 2021-07-15 ms/SI.Rmd
Update MS and SI
Add SI tables, figure. Fix usage of latex formatting in tables.

```{r load-data-si, cache = FALSE}
tar_load(c(
  "japan_shp",
  "comm_scaled_list",
  "occ_point_data_ferns",
  "biodiv_ferns_spatial",
  "biodiv_ferns_endemic_spatial",
  "k_taxonomy",
  "k_phylogeny",
  "traits_nmds",
  "ppgi",
  "phy_sig_results",
  "binary_sig_results",
  "lrt_results",
  "japan_shp"
))
```

@@

```{r phy-sig, results = "as-is"}
phy_sig_results %>%
  mutate(across(c(kval, lambda, k_pval, lambda_pval), ~format(., digits = 2))) %>%
  kbl(booktabs = TRUE) %>%
  kable_styling(latex_options = c("striped", "HOLD_position"))
```

@@

```{r phy-sig-binary, results = "as-is"}
binary_sig_results %>%
  select(trait, D, num_present, num_absent, prob_random, prob_brownian) %>%
  arrange(trait) %>%
  mutate(across(c(D, prob_random, prob_brownian), ~format(., digits = 2))) %>%
  kbl(booktabs = TRUE) %>%
  kable_styling(latex_options = c("striped", "scale_down", "HOLD_position"))
```

**`r s_table("phy-sig-binary")`**. Phylogenetic signal in quantitative (binary) functional traits of the ferns of Japan. D, Fritz and Purvis' *D*; *n* (present), number of times a trait was observed present; *n* (absent), number of times a trait was observed absent; *p* (random), probability of obtaining observed trait distribution given random distribution of traits; *p* (Brownian), probability of obtaining observed trait distribution given random evolution by Brownian Motion. Presence or absence of indusium was originally a binary trait; other traits (margin, shape, texture) are qualitative traits scored in binary format.

@@

```{r lrt-null, results = "as-is"}
lrt_table_data <-
lrt_results %>%
  filter(comparison == "null_model") %>%
  select(-null_formula, -comparison) %>%
  tidyr::extract(full_formula, "model_type", "~ ([^ ]+) ") %>%
  rename_resp_vars %>%
  rename_model_type %>%
  select(model_type, resp_var, chi2_LR, df, loglik_null, loglik_full, p_value) %>%
  arrange(model_type, resp_var) %>%
  mutate(across(where(is.double), ~format(., digits = 2)))

lrt_table_data %>%
  select(-model_type) %>%
  kbl(
    col.names = colnames(.) %>% str_replace_all("resp_var", ""),
    booktabs = TRUE) %>%
  pack_rows(index = base::table(lrt_table_data$model_type)) %>%
  kable_styling(latex_options = c("striped", "HOLD_position"))
```

@@

@@

# Set color palettes
# Randomization test significance
signif_cols <-
  c(
    "> 0.99" = "#e31a1c", 
    "> 0.975" = "#fb9a99",
    "< 0.01" = "#1f78b4", 
    "< 0.025" = "#a6cee3",
    "not significant" = "#ffffcc"
  )

# CANAPE
canape_cols <-
  c(
    "neo" = "#e31a1c", 
    "paleo" = "#1f78b4",
    "mixed" = "#6a51a3",
    "super" = "#4a1486",
    "not significant" = "#ffffcc"
  )
```

@@

a + b + c +
  plot_annotation(tag_levels = 'A')
```

@@

a + b + c + d +
  plot_layout(ncol = 2, nrow = 2) +
  plot_annotation(tag_levels = 'A')
```

@@

a + b +
  plot_layout(ncol = 2) +
  plot_annotation(tag_levels = 'A')
```

@@

>>>>

```{r load-data-si, cache = FALSE}
tar_load(c(
  "japan_shp",
  "comm_scaled_list",
  "occ_point_data_ferns",
  "biodiv_ferns_spatial",
  "biodiv_ferns_endemic_spatial",
  "k_taxonomy",
  "k_phylogeny",
  "traits_nmds",
  "ppgi",
  "phy_sig_results",
  "binary_sig_results",
  "lrt_results",
  "japan_shp",
  "morans_i",
  "t_test_results",
  "aic_env_repro"
))
```

@@

```{r moran, results = "as-is"}
morans_i %>%
  select(resp_var = var, morans_i, pval_i) %>%
  mutate(
    across(c(morans_i, pval_i), ~format(., digits = 2)),
    resp_var = clean_vars(resp_var) %>% 
      factor(levels = c("Area", "% apomictic taxa", "Precipitation", "Precipitation seasonality", "Temperature", "Richness", "SES of PD", "SES of RPD", "SES of FD", "SES of RFD"))
    ) %>%
  assert(not_na, resp_var) %>%
  arrange(resp_var) %>%
  # format for kable/latex: since I use latex in headers, must escape latex elsewhere
  mutate(resp_var = str_replace_all(resp_var, "%", "\\\\%")) %>%
  rename(`\\textit{I}` = morans_i, `\\textit{P}` = pval_i, "Variable" = resp_var) %>%
  kbl(booktabs = TRUE, escape = FALSE, format = "latex") %>%
  kable_styling(latex_options = c("striped", "HOLD_position"))
```

**`r s_table("moran")`**. Spatial autocorrelation in biodiversity metrics (richness, SES of PD, SES of FD, SES of RPD, SES of RFD) and independent variables (environmental variables and % apomictic taxa) as measured with Moran's *I*. Significant *I* > 0 indicates autocorrelation.

```{r corr, results = "as-is"}
t_test_results_for_table <-
t_test_results %>%
  mutate(
    var1 = clean_vars(var1),
    var2 = clean_vars(var2)) %>%
  arrange(var1, var2)

# Identify which rows to make bold based on p < 0.05
t_test_signif <- t_test_results_for_table$p_value < 0.05
t_test_signif[is.na(t_test_signif)] <- FALSE
t_test_signif <- which(t_test_signif)

t_test_results_for_table %>%
  mutate(across(where(is.double), ~format(., digits = 2))) %>%
  # format for kable/latex
  mutate(
    var1 = str_replace_all(var1, "%", "\\\\%"),
    var2 = str_replace_all(var2, "%", "\\\\%")) %>%
  rename(
    "Variable 1" = var1,
    "Variable 2" = var2,
    `\\textit{P}` = p_value,
    `\\textit{r}` = corr,
    `\\textit{f}` = f_stat,
    "df" = dof) %>%
  kbl(booktabs = TRUE, escape = FALSE, format = "latex") %>%
  row_spec(row = t_test_signif, bold = TRUE) %>%
  kable_styling(latex_options = c("striped", "HOLD_position"))
```

**`r s_table("corr")`**. Results of modified *t*-test for correlation between variables while taking into account spatial position. Significantly correlated variables at *P* < 0.05 in bold.

```{r phy-sig, results = "as-is"}
phy_sig_results %>%
  mutate(across(c(kval, lambda, k_pval, lambda_pval), ~format(., digits = 2))) %>%
  # format for kable/latex
  mutate(trait = str_replace_all(trait, "_", "\\\\_")) %>%
  rename(
    "Trait" = trait,
    `\\textit{K}` = kval, `\\textit{P}(\\textit{K})` = k_pval, 
    `λ` = lambda, `\\textit{P}(λ)` = lambda_pval,
    `\\textit{N}` = n) %>%
  kbl(booktabs = TRUE, escape = FALSE, format = "latex") %>%
  kable_styling(latex_options = c("striped", "HOLD_position"))
```

@@

```{r phy-sig-binary, results = "as-is"}
binary_sig_results %>%
  select(trait, D, num_present, num_absent, prob_random, prob_brownian) %>%
  arrange(trait) %>%
  # format for kable/latex
  mutate(trait = str_replace_all(trait, "_", "\\\\_")) %>%
    rename(
    "Trait" = trait,
    `\\textit{N} present` = num_present,
    `\\textit{N} absent` = num_absent,
    `\\textit{D}` = D,
    `\\textit{P}(random)` = prob_random, 
    `\\textit{P}(Brownian)` = prob_brownian
    ) %>%
  kbl(booktabs = TRUE, escape = FALSE, format = "latex") %>%
  kable_styling(latex_options = c("striped", "scale_down", "HOLD_position"))
```

**`r s_table("phy-sig-binary")`**. Phylogenetic signal in quantitative (binary) functional traits of the ferns of Japan. *D*, Fritz and Purvis' *D*; *N* (present), number of times a trait was observed present; *N* (absent), number of times a trait was observed absent; *P*(random), probability of obtaining observed trait distribution given random distribution of traits; *P*(Brownian), probability of obtaining observed trait distribution given random evolution by Brownian Motion. Presence or absence of indusium was originally a binary trait; other traits (margin, shape, texture) are qualitative traits scored in binary format.

@@

```{r lrt-null, results = "as-is"}
# Need to keep the input data separate for pack_rows()
lrt_table_data <-
lrt_results %>%
  filter(comparison == "null_model") %>%
  select(-null_formula, -comparison) %>%
  tidyr::extract(full_formula, "model_type", "~ ([^ ]+) ") %>%
  rename_resp_vars %>%
  rename_model_type %>%
  select(model_type, resp_var, chi2_LR, df, loglik_null, loglik_full, p_value) %>%
  arrange(model_type, resp_var) %>%
  mutate(across(where(is.double), ~format(., digits = 2)))

lrt_table_data %>%
  select(-model_type) %>%
  # format for kable/latex
    rename(
    "Response variable" = resp_var,
    `χ\\textsuperscript{2}` = chi2_LR,
    `LogL(null)` = loglik_null,
    `LogL(full)` = loglik_full,
    `\\textit{P}` = p_value
    ) %>%
  kbl(booktabs = TRUE, escape = FALSE, format = "latex") %>%
  pack_rows(index = base::table(lrt_table_data$model_type)) %>%
  kable_styling(latex_options = c("striped", "HOLD_position"))
```

@@

```{r aic, results = "as-is"}
aic_table_data <- aic_env_repro %>%
  rename_resp_vars %>%
  rename_model_type %>%
  select(model_type, resp_var, conditional_aic, df = effective_df, loglik) %>%
  arrange(model_type, resp_var) %>%
  mutate(across(where(is.double), ~format(., digits = 2)))

aic_table_data %>%
  select(-model_type) %>%
  # format for kable/latex
    rename(
    "Response variable" = resp_var,
    "cAIC" = conditional_aic,
    "LogL" = loglik
    ) %>%
  kbl(booktabs = TRUE, escape = FALSE, format = "latex") %>%
  pack_rows(index = base::table(aic_table_data$model_type)) %>%
  kable_styling(latex_options = c("striped", "HOLD_position"))
```

**`r s_table("aic")`**. Model fit as measured with conditional Akaike Information Criterion (cAIC). Environmental models included the effects of temperature, precipitation, precipitation seasonality, and area. Reproductive models included the effects of % apomictic taxa, precipitation, precipitation seasonality, and area. Lower cAIC indicates better fit.

@@

# Set color palettes
# - Randomization test significance
signif_cols <-
  c(
    "> 0.99" = "#e31a1c", 
    "> 0.975" = "#fb9a99",
    "< 0.01" = "#1f78b4", 
    "< 0.025" = "#a6cee3",
    "not significant" = "#ffffcc"
  )
# - CANAPE
canape_cols <-
  c(
    "neo" = "#e31a1c", 
    "paleo" = "#1f78b4",
    "mixed" = "#6a51a3",
    "super" = "#4a1486",
    "not significant" = "#ffffcc"
  )
# - Bioregions
bioregion_cols <- rcartocolor::carto_pal(9, "Bold")[1:8] %>%
  set_names(1:8)
```

@@

a + b + c +
  plot_annotation(tag_levels = "A") &
    theme(plot.tag = element_text(face = "bold"))
```

@@

a + b + c + d +
  plot_layout(ncol = 2, nrow = 2) +
  plot_annotation(tag_levels = "A") &
    theme(plot.tag = element_text(face = "bold"))
```

@@

a + b +
  plot_layout(ncol = 2) +
  plot_annotation(tag_levels = "A") &
    theme(plot.tag = element_text(face = "bold"))
```

@@

```{r bioregions-env, fig.width = 7, fig.height = 5}
# Make pot showing bioregions by temperature and precipitation
a <-
ggplot(biodiv_ferns_spatial, aes(x = precip, y = temp, color = taxonomic_cluster)) +
  geom_point(size = 1, alpha = 0.7) +
  scale_color_manual(values = bioregion_cols) +
  scale_y_continuous(labels = function(x) x/10) +
  labs(y = "Temperature (°C)", x = "Precipitation (mm)", color = "Cluster", subtitle = "Taxonomic bioregions") +
  theme(legend.position = "bottom")

b <-
  ggplot(biodiv_ferns_spatial, aes(x = precip, y = temp, color = phylo_cluster)) +
  geom_point(size = 1, alpha = 0.7) +
  scale_color_manual(values = bioregion_cols) +
  scale_y_continuous(labels = function(x) x/10) +
  labs(y = "Temperature (°C)", x = "Precipitation (mm)", color = "Cluster", subtitle = "Phylogenetic bioregions") +
  theme(legend.position = "bottom")

a + b + plot_annotation(tag_levels = "A") &
    theme(plot.tag = element_text(face = "bold"))
```

**`r s_figure("bioregions-env")`**. Scatterplots of grid cell membership in (**A**) taxonomic or (**B**) phylogenetic bioregions arranged by mean annual temperature and annual precipitation. Points shown with partial transparency; darker areas indicate overlapping points.

### 2021-07-15 ms/manuscript.Rmd
Update MS and SI
Add SI tables, figure. Fix usage of latex formatting in tables.

Joel H. Nitta^1^, Brent D. Mishler^2^, Wataru Iwasaki^1, 3, 4, 5, 6^, and Atsushi Ebihara^7^

@@

All measures of biodiversity and CANAPE were carried out using the `r pack("canaper")` package (https://github.com/joelnitta/canaper), which calculates raw biodiversity metrics using functions in the `r pack("phyloregion")` package [@Daru2020a] and randomizations using the `r pack("picante")` package [@Kembel2010].

@@

We first tested for spatial autocorrelation in biodiversity metrics (richness, SES of PD, SES of FD, SES of RPD, SES of RFD) and independent variables (environmental variables and % apomictic taxa) using Moran's I with the `r func("moran.mc")` function in the `r pack("spdep")` package. 
In all cases, significant spatial autocorrelation was detected (Moran's I > 0, *P* value of permutation test < 0.05). 
Therefore, we used spatially explicit methods for modeling.

We checked for correlation between independent variables while taking into account spatial autocorrelation using a modified *t*-test with the `r func("modified.ttest")` function in the `r pack("SpatialPack")` R package. 
This indicated that temperature seasonality and % apomictic taxa were correlated with mean annual temperature, so for the main modeling analysis we only included the following environmental variables: mean annual temperature, precipitation, and precipitation seasonality.

@@

Code to replicate all analyses, figures, and this manuscript are available at https://github.com/joelnitta/japan_ferns_biogeo. 
Raw and processed data are openly available from the Dryad repository at http://doi.org/ADD_DOI_HERE_WHEN_ASSIGNED.

@@

All models were significant relative to null models without any fixed effects `r s_table("lrt-null")`. 
The effect of temperature dominated (had the greatest absolute *t*-value) in environmental models (`r figure("coeff")`A--E), and % apomictic taxa dominated in reproductive models (`r figure("coeff")`F--G); temperature and % apomictic taxa were significant in each model where they were included, whereas other predictors (area, precipitation, precipitation seasonality) were not significant in some models. 
Richness showed a hump-shaped relationship with temperature, reaching maximum richness at intermediate temperatures (`r figure("modelfit")`A). 
Phylogenetic and functional diversity showed either moderately (SES of PD, SES of FD) to strongly (SES of RPD, SES of RFD) hump-shaped relationships with temperature (`r figure("modelfit")`B--E). 
SES of RPD showed a clear, linear decline with % apomictic taxa (`r figure("modelfit")`G); a similar pattern was detected in SES of PD, but with wider confidence intervals (`r figure("modelfit")`F). 
Conditional AIC was lower (indicating better model fit) in the reproductive model compared to the environmental model for SES of RPD, but not SES of PD.

@@

The variation in distribution of different biodiversity metrics is reflected by the bioregions analysis, which tended to group cells with similar biodiversity values (`r figure("div-by-tax-region")`). 
Interestingly, the bioregions identified here correspond roughly to major forest types previously categorized on an *ad-hoc* basis; Cluster 1 corresponds to the cool or subartic deciduous type, Cluster 2 to the warm evergreen type, and Clusters 3 and 4 to the subtropical rainy type [@Shimizu2014]. 
This suggests that ferns are useful as possible bioindicators in Japan. 
Although some studies have emphasized the importance of deep-sea straits (the Tsugaru Strait separating Hokkaido from the rest of Honshu and the Tokara Strait separating the southern islands from Kyushu) in structuring biological communities in Japan [@Millien-Parra1999; @Kubota2014], these did not play a major role in structuring fern bioregions. 
The Tsugaru Strait had no relation to any of the bioregions, and although the Tokara Strait somewhat splits clusters 2 and 3, there is some overlap between them on the Amami-Oshima Isl (`r figure("bioregions")`). 
The weak effect of these straits as barriers makes sense given the presumably high dispersal abilities of ferns, which disperse via tiny spores carried on the wind.

>>>>

Joel H. Nitta^1^, Brent D. Mishler^2^, Wataru Iwasaki^1,3,4,5,6^, and Atsushi Ebihara^7^

@@

All measures of biodiversity and CANAPE were carried out using the `r pack("canaper")` package [@Nitta2021], which calculates raw biodiversity metrics using functions in the `r pack("phyloregion")` package [@Daru2020a] and randomizations using the `r pack("picante")` package [@Kembel2010].

@@

We first tested for spatial autocorrelation in biodiversity metrics (richness, SES of PD, SES of FD, SES of RPD, SES of RFD) and independent variables (environmental variables and % apomictic taxa) using Moran's I with the `r func("moran.mc")` function in the `r pack("spdep")` package. 
In all cases, significant spatial autocorrelation was detected (Moran's I > 0, *P* value of permutation test < 0.05) (`r s_table("moran")`). 
Therefore, we used spatially explicit methods for modeling.

We checked for correlation between independent variables while taking into account spatial autocorrelation using a modified *t*-test with the `r func("modified.ttest")` function in the `r pack("SpatialPack")` R package. 
This indicated that temperature seasonality and % apomictic taxa were correlated with mean annual temperature (`r s_table("corr")`), so for the main modeling analysis we only included the following environmental variables: mean annual temperature, precipitation, and precipitation seasonality.

@@

Code to replicate all analyses, figures, and this manuscript are available at https://github.com/joelnitta/japan_ferns_biogeo. 
A Docker image to run the code is available at https://hub.docker.com/repository/docker/joelnitta/japan_ferns_biogeo.
Data are openly available from Figshare at http://doi.org/ADD_DOI_HERE_WHEN_ASSIGNED.

@@

All models were significant relative to null models without any fixed effects `r s_table("lrt-null")`. 
The effect of temperature dominated (had the greatest absolute *t*-value) in environmental models (`r figure("coeff")`A--E), and % apomictic taxa dominated in reproductive models (`r figure("coeff")`F--G); temperature and % apomictic taxa were significant in each model where they were included, whereas other predictors (area, precipitation, precipitation seasonality) were not significant in some models. 
Richness showed a hump-shaped relationship with temperature, reaching maximum richness at intermediate temperatures (`r figure("modelfit")`A). 
Phylogenetic and functional diversity showed either moderately (SES of PD, SES of FD) to strongly (SES of RPD, SES of RFD) hump-shaped relationships with temperature (`r figure("modelfit")`B--E). 
SES of RPD showed a clear, linear decline with % apomictic taxa (`r figure("modelfit")`G); a similar pattern was detected in SES of PD, but with wider confidence intervals (`r figure("modelfit")`F). 
Conditional AIC was lower (indicating better model fit) in the reproductive model compared to the environmental model for SES of RPD, but not SES of PD (`r s_table("aic")`).

@@

The variation in distribution of different biodiversity metrics is reflected by the bioregions analysis, which tended to group cells with similar biodiversity values (`r figure("div-by-tax-region")`). 
Interestingly, the bioregions identified here correspond roughly to major forest types previously categorized on an *ad-hoc* basis; Cluster 1 corresponds to the cool or subartic deciduous type, Cluster 2 to the warm evergreen type, and Clusters 3 and 4 to the subtropical rainy type [@Shimizu2014]. 
This suggests that ferns are useful as possible bioindicators in Japan. 
Although some studies have emphasized the importance of deep-sea straits (the Tsugaru Strait separating Hokkaido from the rest of Honshu and the Tokara Strait separating the southern islands from Kyushu) in structuring biological communities in Japan [@Millien-Parra1999; @Kubota2014], these did not play a major role in structuring fern bioregions. 
The Tsugaru Strait had no relation to any of the bioregions, and although the Tokara Strait somewhat splits clusters 2 and 3, there is some overlap between them on the Amami-Oshima Isl (`r figure("bioregions")`). 
The weak effect of these straits as barriers makes sense given the presumably high dispersal abilities of ferns, which disperse via tiny spores carried on the wind.
Rather, fern bioregions in Japan broadly separate along gradients of temperature and precipitation; in particular, precipitation splits taxonomic Cluster 3 (mostly the Ryukyu islands, wetter) from Cluster 4 (mostly the Ogasawara islands, drier) (`r s_figure("bioregions-env")`).
